---
title: "Week 8 Lab"
format: html
---


The following materials are based on materials developed by Dr. Chris Kibler for the UCSB Geography Department.

## Background

Phenology is the timing of life history events. Important phenological events for plants involve the growth of leaves, flowering, and senescence (death of leaves). Plants species adapt the timing of these events to local climate conditions to ensure successful reproduction. Subsequently, animal species often adapt their phenology to take advantage of food availability. As the climate shifts this synchronization is being thrown out of whack. Shifts in phenology are therefore a common yardstick of understanding how and if ecosystems are adjusting to climate change.

Plant species may employ the following phenological strategies:

- Winter deciduous: lose leaves in the winter, grow new leaves in the spring
- Drought deciduous: lose leaves in the summer when water is limited
- Evergreen: maintain leaves year-round


## Task

In this lab we are analyzing plant phenology near the Santa Clara River which flows from Santa Clarita to Ventura. We will investigate the phenology of the following plant communities:

- Riparian forests: grow along the river, dominated by winter deciduous cottonwood and willow trees
- Grasslands: grow in openspaces, dominated by drought deciduous grasses
- Chaparral shrublands: grow in more arid habitats, dominated by evergreen shrubs

To investigate the phenology of these plant communities we will use a time series of Landsat imagery and polygons identifying the locations of study sites within each plant community.

Our primary goal is to compare seasonal patterns across vegetation communities. To do so, we will:

- Convert spectral reflectance into a measure of vegetation productivity (NDVI)
- Calculate NDVI throughout the year
- Summarize NDVI values within vegetation communities
- Visualize changes in NDVI within vegetation communities


## Set Up 
```{r}
library(terra)
library(sf)
library(tidyverse)
library(here)
library(tmap)
```

## Create function to compute NDVI
```{r}
ndvi_fun <- function(nir, red){
  (nir - red)/(nir + red)
}
```

## Compute NDVI for a single scene 

```{r}
# Load in data for June 2018
landsat_20180612 <- terra::rast(here("data", "landsat_20180612.tif"))

# Rename layers (bands)
names(landsat_20180612) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")

# Use lappy function to compute NDVI
ndvi_20180612 <- lapp(landsat_20180612[[c(4,3)]], fun = ndvi_fun)

# Make a map to show NDVI
tm_shape(ndvi_20180612) +
  tm_raster(title = "NDVI") +
  tm_layout(legend.outside = TRUE)
```

### Compute NDVI for all scenes 

Create a function that will do what we did above for all the scenes

```{r}
# Clear environment
rm(list = ls())

# Function again 
ndvi_fun <- function(nir, red){
  (nir - red)/(nir + red)
}
```

```{r}
# Create list of file names 
files <- list.files(here("data"), # where the files are located
                    pattern = "*.tif", # files that end with .tif
                    full.names = TRUE) # gives the full path


create_ndvi_layer <- function(i){
  #step 1: read in scene
  landsat <- rast(files[i])
  
  #step 2: rename bands in scene
  names(landsat) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")
  
  #step 3: compute NDVI
  ndvi <- lapp(landsat[[c(4,3)]], fun = ndvi_fun)
  
  return(ndvi)
}
```

```{r}
# Read in all files to one raster stack using function 
all_ndvi <- c(create_ndvi_layer(1),
              create_ndvi_layer(2),
              create_ndvi_layer(3),
              create_ndvi_layer(4),
              create_ndvi_layer(5),
              create_ndvi_layer(6),
              create_ndvi_layer(7),
              create_ndvi_layer(8))

# Update names 
names(all_ndvi) <- c("2018-06-12", 
                    "2018-08-15", 
                    "2018-10-18", 
                    "2018-11-03", 
                    "2019-01-22", 
                    "2019-02-23", 
                    "2019-04-12", 
                    "2019-07-01")
```

Can we do this more effectively
```{r}
# Extract the date from the file name to rename the layers 
file_names <- data.frame(file_name = files) %>% 
  mutate(date = str_sub(file_name, start = -12L, end = -5L)) # - = from the end of file name

names(all_ndvi) <- file_names$date
```

## Compare NDVI across veg communities 

### Get data on veg communities 

```{r}
sites <- st_read(here("data", "study_sites.shp"))

tm_shape(all_ndvi[[1]]) +
  tm_raster() +
  tm_shape(sites) +
  tm_borders()
```

### Extract NDVI at study sites 
```{r}
sites_ndvi <- terra::extract(all_ndvi, sites, fun = "mean")

# Bind it back to og df to keep information 
sites_annotated <- cbind(sites, sites_ndvi)
```

We’re done! Except our data is very untidy… Let’s tidy it up!

- Convert to data frame
- Turn from wide to long format
- Turn layer names into date format
- Combine study sites by vegetation type
- Summarize results within vegetation types

```{r}
sites_clean <- sites_annotated %>% 
  # drop columns
  select(-ID) %>% 
  st_drop_geometry() %>% 
  # reformat data frame to long
  pivot_longer(!study_site) %>% 
  rename("NDVI" = value) %>% 
  # create date 
  mutate("year" = str_sub(name, 2, 5),
         "month" = str_sub(name, 6, 7),
         "day" = str_sub(name, 8, 9)) %>% 
  # combine date elements 
  unite("date", 4:6, sep = "-") %>% # collapse y m d into one column
  mutate("date" = lubridate::as_date(date)) %>% 
  # remove name column
  select(-name) %>% 
  # create new column to collapse forest 1-3
  mutate("veg_type"= case_when(study_site== "forest1" ~ "forest",
                               study_site== "forest2" ~ "forest",
                               study_site== "forest3" ~ "forest",
                               study_site== "grassland" ~ "grassland",
                               study_site== "chaparral"~"chaparral")) %>% 
  group_by(veg_type, date) %>% 
  summarise("NDVI" = mean(NDVI, na.rm = TRUE))
```

```{r}
ggplot(sites_clean, 
       aes(x = date, y = NDVI, col = veg_type)) +
  geom_line() +
  geom_point() +
  theme_light() + 
  labs(x = " ", 
       y = "Normalized Difference Vegetation Index (NDVI",
       col = "Vegetation Type", 
       title = "Seasonal cycle of vegetation productivity")
```

